\documentclass[letterpaper, 12pt]{article}
\usepackage{amsmath, amssymb, mathtools} % essentials
\usepackage[left=1in,right=1in,top=1in,bottom=1.5in]{geometry}
\renewcommand{\baselinestretch}{1.1}
\usepackage{enumerate}
\usepackage{float}
\usepackage{amsthm}
\newtheorem{thm}{Theorem}
\newtheorem{conj}{Conjecture}
%\usepackage{tikz-cd}
%\usepackage{xcolor}
%\usepackage{url}
%\usepackage{graphicx}
%\graphicspath{{./}}
%\usepackage{enumitem}
%\usepackage{comment}

% convenience names
\newcommand{\nequiv}{\not\equiv}
\newcommand{\nsubset}{\not\subset}
\renewcommand{\emptyset}{\varnothing}
\newcommand{\xto}{\xrightarrow}
\newcommand{\injto}{\hookrightarrow}
\newcommand{\surjto}{\twoheadrightarrow}

\newcommand{\diff}{\mathop{}\!d} % as in $\int e^x \diff x$

\renewcommand{\Re}{\operatorname{Re}}
\renewcommand{\Im}{\operatorname{Im}}
\DeclareMathOperator{\Cl}{Cl} % ideal class group
\DeclareMathOperator{\Frac}{Frac}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Gal}{Gal}
\DeclareMathOperator{\Stab}{Stab}

% code from https://tex.stackexchange.com/a/2610/223981 that fixes a spacing issue with \left\right
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}

\title{15-Theorem and Its Generalizations: A Journey from Ramanujan to Conway and Bhargava\footnote{Fall 2024 Math 254A Term Paper. I wrote this myself, with help from Google Gemini AI for generating citations.}: Presentation Edition}
\author{Mingyang Cen \\ Student ID: 3036462810 \\ Email address: \texttt{mcen@berkeley.edu}}

\begin{document}

\maketitle

\section{Introduction}
Lagrange's famous four-square theorem states that the quadratic form $x^2+y^2+z^2+w^2$ represents all natural numbers. One may wonder which other quadratic forms have this property of representing all natural numbers, which we will call ``universality.'' A solution is provided by the 15-Theorem and 290-Theorem, which I will discuss in this talk.

First, we need to specify which quadratic forms we are looking at. For a quadratic form, ``integer-matrix'' means that when the quadratic form is represented as a symmetric matrix, all entries are integers. ``Integer-valued'' is a more lenient condition that allows off-diagonal entries to be half-integers.

The 15-Theorem states that a positive-definite integer-matrix quadratic form (which I will refer to as simply a quadratic form from now on) represents all positive integers exactly when it represents 1, 2, 3, 5, 6, 7, 10, 14, and 15. These nine numbers are called critical numbers.

The number 15 is special in this theorem since, in fact,
a quadratic form representing all critical numbers besides $15$ must represent all positive integers besides $15$.
Also, by application of the 15-Theorem, we easily obtain an extremely useful classification of universal four-dimensional quadratic forms, namely, that there are exactly $204$ non-isomorphic universal four-dimensional quadratic forms.

The 290-Theorem states the same thing as the 15-Theorem, except we allow integer-valued quadratic forms, and the critical numbers are enlarged to a set of 29 numbers, up to 290.
The 290-Theorem appears to be similar to the 15-Theorem, and the proof has the same structure. However, the weakened condition of being integer-valued vastly increases the number of cases that must be considered. Because of this, the 290-Theorem requires several analytic results on modular forms for the proof, so I will not present it today.
There is a similar classification of universal four-dimensional integer-valued quadratic forms.

I have created SageMath code for some computations required for the 15-theorem.

\section{Definitions}
There is a well-known natural bijection between lattices with integer inner products and positive-definite integer-matrix quadratic forms, so any operation on a lattice can be applied to a quadratic form, and vice versa. I am going to use lattices and quadratic forms in the proof, whichever is most convenient.

I will present several important pieces of machinery for the proof. A quadratic form or its corresponding lattice is defined to be \emph{universal} if it represents all positive integers. If it is not universal, its \emph{truant} is defined to be the smallest positive integer not represented. An \emph{escalation} of a lattice $L$ is a lattice generated by $L$ and a vector of norm equal to the truant of $L$. An \emph{escalator lattice} is one reached from a series of escalations from the zero lattice.

\section{Proof of the 15-Theorem}
\subsection{Escalators of dimension 1, 2, and 3}
Through computation, there are 9 three-dimensional escalators.
Escalating again gives 207 nonisomorphic four-dimensional escalator lattices. An important issue is to determine that the quaternary escalators either have truant in the critical numbers or are universal.

\subsection{Escalators of dimension 4}
We want to show 201 of the 207 quaternary escalators are universal, and the other 6 have truants 10 or 15.

As an illustrative example for the argument for the 201 universal quaternaries, let $L$ be an escalation of $L' = \begin{bsmallmatrix} 1 & 0 & 0 \\ 0 & 2 & 0 \\ 0 & 0 & 2 \end{bsmallmatrix}$, where the orthogonal complement of $L'$ in $L$ has Gram matrix with one entry $m$. Since $L'$ is unique in its genus, we may apply the Hasse-Minkowski principle to go from the global setting of the integers to the local setting of the $p$-adic numbers. Through some application of Hensel's lemma, we find that $L'$ represents everything but positive integers of the form $4^a (8b + 7)$, where $a$ and $b$ are nonnegative integers.

Suppose for the sake of contradiction that $L$ is not universal, and let its truant be $u$. Since $L'$ cannot represent $u$, we can write $u$ in the form $4^a (8b + 7)$. Since $u$ is minimal, it is squarefree, so $a = 0$ and $u \equiv 7 \pmod 8$. We consider three cases depending on $m$ modulo $8$:
\begin{itemize}
    \item If $m$ is not congruent to $0$, $3$, or $7$ modulo $8$, then $u-m$ cannot be of the form $4^a (8b + 7)$.
    \item If $m \equiv 3, 7 \pmod 8$, then $u-4m$ cannot be of the form $4^a (8b + 7)$ either.
    \item If $m \equiv 0 \pmod 8$, then $L$ is called exceptional, and we have to use a different $L'$.
\end{itemize}

Therefore, either $u < 4m$ or $u - m$ or $u - 4m$ (depending on $m \pmod 8$) is represented by $L'$. Because $L = L' \oplus [m]$, if $u - m$ or $u - 4m$ is represented by $L'$, $u$ is represented by $L$. This is false by assumption. Hence, $u$ must be less than $4m$. By computation, $m \le 28$, so $u < 112 = 4 \cdot 28$. Now that we have bounded $u$, we use computer checks to show that $L$ represents all positive integers up to $112$. This shows $u$ did not exist in the first place, so $L$ is universal.

There are a few special cases where a slightly modified argument must be applied, but in the end, 201 quaternary escalators are shown to be universal.

\subsection{Escalators of dimension 5}
The $6$ nonuniversal four-dimensional escalators have truants $10,10,15,15,15,15$. It is impractical to consider every single one of the 1630 five-dimensional escalators, but we have a trick. We may apply the same argument as we just did for the other 201 four-dimensional escalators, but the final check that $L$ represents all positive integers up to some number fails with just one number. This shows that these $6$ lattices each miss that number and no others. That number is the lattice's truant, so the $5$-dimensional escalators of the lattice must be universal. This finishes our analysis of escalators.

For every quadratic form $Q$ representing each critical number, there exists an escalator within $Q$ representing every critical number. Through analysis of the escalators, we find that the escalator within $Q$ is in fact universal. Hence, $Q$ is universal, proving the 15-theorem.

The 15-Theorem also provides a shortcut to classifying all universal $4$-variable positive-definite integer-matrix quadratic forms. We find a bound on the discriminant of such a quadratic form and systematically apply the $15$-Theorem in each case to find there are exactly $204$ non-isomorphic universal quaternaries.

\section{Related results}

Several generalizations of the 15-Theorem and the 290-Theorem exist: to quadratic forms representing odd numbers, prime numbers, numbers coprime to a certain number, etc.

\end{document}
